Grove: A Convergent Collaborative Structure-Editor Calculus

### 2020-11-18 grove-paper.tex
comment out Eric's notes appendix
Already folded in useful parts. All that's left are inter-editor
semantics, which we're not covering here.

\section{Appendix}
\label{apx:Appendix}

\subsection{Concepts}

\begin{tabular}{cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl}
  $\A$ & Actions      & $\S$  & Edge States   \\
  %  $\C$ & Cursors & $\U$ & UUIDs         \\
  $\E$ & Edges        & $\V$  & Vertexes      \\
  $\G$ & Graphs       & $\AA$ & Graph Actions \\
  $\I$ & Indices      & $\EE$ & Editors       \\
  $\K$ & Constructors                         \\
\end{tabular}

A \emph{UUID} $u \in \U$ is a universally unique identifier.

\subsubsection{The Target Language}

\paragraph{Constructor} \par

A \emph{constructor} $k \in \K$ represents a node in the target language's abstract syntax tree (AST). Let $k_{root}$ be the root AST node.

\paragraph{Index} \par

An \emph{index} $i \in \I$ identifies a position, relative to some parent constructor, of a potential child node in the target language's AST. Let $i_{root}$ be the hyper-position of all rooted constructors.

\subsubsection{The Graph}

\paragraph{Vertex} \par

A \emph{vertex} $v = (k, u) \in \V = \K \times \U$ is a distinct instance of constructor $k$; that is, one that can be identified by UUID $u$. Let $v_{root} = (k_{root}, u_{root})$ be the root vertex of a graph. The \emph{rooted vertexes} of a graph are the immediate children of its $v_{root}$.

\paragraph{Cursor} \par

%A \emph{cursor} $c = (v,i) \in \C = \V \times \I$ is a reference to the (possibly empty) set of all vertexes with parent vertex $v$ and child index $i$. Let $c_{root} = (v_{root}, i_{root})$ be the \emph{root cursor}, a reference to all of the rooted vertexes, of a graph.

\paragraph{Edge} \par

%An \emph{edge} $\e = (c, v, u) \in \E = \C \times \V \times \U$ adds vertex $v$ with the set of vertexes referenced by cursor $c$.

\paragraph{Edge State} \par

An \emph{edge state} $s \in \S = \{\bot,+,-\}$ determines whether an edge does not exist ($\bot$), has been created (+), or has been destroyed (-).

\paragraph{Graph Action} \par

A \emph{graph action} $A = (\e,s,u) \in \A = \E \times \S \times \U$ is an instance of a binding from edge $\e$ to edge state $s$ that can be identified by UUID $u$. Denote by $[\e :-> s]_{u}$ a binding from edge $\e$ to edge state $s$ (identified by UUID $u$).

\paragraph{Graph} \par

A \emph{graph} $G = [\e :-> s]_{u}^{\ast{}} \in \G = \A^{\ast{}}$ is a function from edges to edge states. Denote by $G(\e) = s$ the state $s$ of edge $\e$ in graph $G$, and by $G A$ the extension of graph $G$ by graph action $A$.

\subsubsection{The Editor}

\paragraph{Move Action} \par

%A \emph{move action} $a \in \A^{move} = \{ \Left,\Right,\Up,\Down \} \cup \{ \Select \} \times \C$ re-positions the cursor.

\paragraph{Edit Action} \par

An \emph{edit action} $a \in \A^{edit}= \{ \Create \} \times \K \cup \{ \Destroy \} \cup \{ \Restore \} \times \V$ adds or removes an edge at the cursor.

\paragraph{Communication Action} \par

A \emph{communication action} $a = (\Send, A^{\ast{}}, u^{\ast{}}) \in \A^{comm} = \{ \Send \} \times \A^{\ast{}} \times \U^{\ast{}}$ applies the graph action sequence $A^{\ast{}}$ to the editors identified by UUIDs $u^{\ast{}}$.

\paragraph{Action} \par

An \emph{action} $a \in \A = \A^{move} \cup \A^{edit} \cup \A^{comm}$ describes a change to one or more editors.

\paragraph{Editor} \par

%An \emph{editor} $E = (G,c,A_{Q}^{\ast{}},A_{H}^{\ast{}},u) \in \EE = \mathcal{G} \times \C \times \AA^{\ast{}} \times \AA^{\ast{}} \times \U$ is an instance of graph $G$ with cursor $c$, queued graph actions $A_{Q}^{\ast{}}$, and known graph action history $A_{H}^{\ast{}}$, that can be identified by UUID $u$.

\subsubsection{The Environment} \par

An \emph{environment} $(E^{\ast{}}, A^{\ast{}}) \in \EE^{\ast{}} \times \AA^{\ast{}}$ is a set of communicating editors $E^{\ast{}}$ that share a global graph action history $A^{\ast{}}$.

\subsection{Sets and Relations}

\paragraph{Edges} \par

\begin{align*}
  \edges(G)            & = \dom(G)                                       \\
  \liveEdges(G)        & = \{ \e \in \edges(G) | G(\e)=+ \}              \\
  \vertexChildren(G,v) & = \{ ((v',i),v'',u) \in \liveEdges(G) | v'=v \} \\
  \cursorChildren(G,c) & = \{ (c',v,u) \in \liveEdges(G) | c'=c \}       \\
  \parents(G,v)        & = \{ (c,v',u) \in \liveEdges(G) | v'=v \}
\end{align*}

\paragraph{Vertexes} \par

\begin{align*}
  \vertexes(G)        & = \{ v \in \V | \exists (c,v,u) \in \edges(G) \}              \\
  \liveVertexes(G)    & = \{ v \in \V | \exists (c,v,u) \in \liveEdges(G) \}          \\
  \childVertexes(G,v) & = \{ v' \in \V | \exists (c,v',u) \in \vertexChildren(G,v) \} \\
  \orphans(G)         & = \{ v \in \liveVertexes(G) | \abs{\parents(G,v)} = 0 \}      \\
  \multiparents(G)    & = \{ v \in \liveVertexes(G) | \abs{\parents(G,v)} > 1 \}      \\
  \vertex(G,u)        & = v \in \vertexes(G) \iff (\exists k \in \K)(v = (k,u))
\end{align*}

\paragraph{Graph Actions} \par

$\known(E_1 \cdots E_n) = A_{H1}^{\ast{}} \cdots A_{Hn}^{\ast{}}
  \iff (\forall j=1,\ldots,n)(E_j = (G_j,c_j,A_{Qj}^{\ast{}},A_{Hj}^{\ast{}},u_j))$

\section{Operational Semantics}

\subsection{Movement}

$\boxed{E,A -->a E,A}$

\begin{mathpar}
  \inferrule{
  \leftIndex(i) = i'
  }{
  (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Left} (G,(v,i'),A_Q^{\ast{}},A_H^{\ast{}},u),A
  }

  \inferrule{
  \rightIndex(i) = i'
  }{
  (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Right} (G,(v,i'),A_Q^{\ast{}},A_H^{\ast{}},u),A
  }

  \inferrule{
  \parents(v) = \{(c',v',u')\}
  }{
  (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Up} (G,c',A_Q^{\ast{}},A_H^{\ast{}},u),A
  }

  \inferrule{
  \cursorChildren(c) = \{(c',(k,u_k),u')\} \\
  \downIndex(k) = i
  }{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Down} (G,((k,u_k),i),A_Q^{\ast{}},A_H^{\ast{}},u),A
  }

  \inferrule{}{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Select~c'} (G,c',A_Q^{\ast{}},A_H^{\ast{}},u),A
  }
\end{mathpar}

\subsection{Editing}

\begin{mathpar}
  \inferrule{
  \defaultIndex(c) = TODO(Was None) \\
  u, u_k \in \U \text{ fresh} \\
  A_H' = [(c,(k,u_k),u) :-> +]
  }{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Create~k} (G A_H',c,A_Q^{\ast{}},A_H^{\ast{}}A_H',u),A
  }

  \inferrule{
  \defaultIndex(c) = i' \\
  c' = ((k,u_k),i') \\
  u, u_k \in \U \text{ fresh} \\
  \\\\
  \cursorChildren(G,c) = \{(c_j,v_j,u_j')\}_{j=1}^n \\
  A_H'^{\ast{}} =
    [(c  ,(k,u_k),u  ) :-> +]
    [(c' ,v_j    ,u_j') :-> +]_{j=1}^n
  [(c_j,v_j    ,u_j') :-> -]_{j=1}^n
  }{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Create~k}
  (G A_H'^{\ast{}},c',A_Q^{\ast{}},A_H^{\ast{}}A_H'^{\ast{}},u),A
  }

  \inferrule{
  \cursorChildren(G,c) = \{\e_j\}_{j=1}^n \\
  A_H'^{\ast{}} = [\e_j :-> -]_{j=1}^n
  }{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Destroy}
  (G A_H'^{\ast{}},c,A_Q^{\ast{}},A_H^{\ast{}}A_H'^{\ast{}},u),A
  }

  \inferrule{
  A_H = [(c,v,u') :-> +] \\
  u' \text{ fresh}
  }{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Restore~v} (G A_H,c,A_Q^{\ast{}},A_H^{\ast{}}A_H,u),A %
  }
\end{mathpar}

\subsection{Communication}

\paragraph{Record} \par
%
\begin{mathpar}
  \inferrule{
  a \in \A^{edit} \\
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A -->a (G',c',A_Q^{\ast{}},A_H'^{\ast{}},u),A
  }{
  (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A -->a (G',c',A_Q^{\ast{}}a,A_H'^{\ast{}},u),A
  }
\end{mathpar}

\paragraph{Replay} \par

$\boxed{E,A \xrightarrow{a^{\ast{}}} E,A}$
%
\begin{mathpar}
  \inferrule{
    E,A -->a E'',A'' \\
    E'',A'' \xrightarrow{a^{\ast{}}} E',A' \\
  }{
    E,A \xrightarrow{aa^{\ast{}}} E',A'
  }
\end{mathpar}

\paragraph{Publish} \par

$\boxed{(E,A)^{\ast{}} -->a (E,A)^{\ast{}}}$
%
\begin{mathpar}
  \inferrule{
  E_j = (G_j,c_j,A_{Qj}^{\ast{}},A_{Hj}^{\ast{}},u_j) \\
  E_j,A_j \xrightarrow{a'^{\ast{}}} (G_j',c_j',A_{Qj}'^{\ast{}}, A_{Hj}'^{\ast{}},u_j),A_j' \\
  E_j' =
  ( G_j'
  , c_j'
  , A_{Qj}'^{\ast{}} \setminus a'^{\ast{}}
  , A_{Hj}'^{\ast{}} \setminus \known(E_1 \cdots E_n),u_j) \\
  j = 1,\ldots,n
  }{
  (E_1,A_1) \cdots (E_n,A_n) \xrightarrow{\Send~ a'^{\ast{}} } (E_1',A_1') \cdots (E_n',A_n')
  }
\end{mathpar}

Missing Action.env members: Record / Report / Stop / Replay, Dump / Load, Clone / Drop

$G \overset{\left(e,s\right)}{\longrightarrow} G\left[ e \mapsto s \sqcup G(e) \right]$

% Edge.Map.add action.value.edge (join old_state new_state) graph

>>>>

% \section{Appendix}
% \label{apx:Appendix}

% \subsection{Concepts}

% \begin{tabular}{cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl@{\hspace{1.5cm}}cl}
%   $\A$ & Actions      & $\S$  & Edge States   \\
%   %  $\C$ & Cursors & $\U$ & UUIDs         \\
%   $\E$ & Edges        & $\V$  & Vertexes      \\
%   $\G$ & Graphs       & $\AA$ & Graph Actions \\
%   $\I$ & Indices      & $\EE$ & Editors       \\
%   $\K$ & Constructors                         \\
% \end{tabular}

% A \emph{UUID} $u \in \U$ is a universally unique identifier.

% \subsubsection{The Target Language}

% \paragraph{Constructor} \par

% A \emph{constructor} $k \in \K$ represents a node in the target language's abstract syntax tree (AST). Let $k_{root}$ be the root AST node.

% \paragraph{Index} \par

% An \emph{index} $i \in \I$ identifies a position, relative to some parent constructor, of a potential child node in the target language's AST. Let $i_{root}$ be the hyper-position of all rooted constructors.

% \subsubsection{The Graph}

% \paragraph{Vertex} \par

% A \emph{vertex} $v = (k, u) \in \V = \K \times \U$ is a distinct instance of constructor $k$; that is, one that can be identified by UUID $u$. Let $v_{root} = (k_{root}, u_{root})$ be the root vertex of a graph. The \emph{rooted vertexes} of a graph are the immediate children of its $v_{root}$.

% \paragraph{Cursor} \par

% %A \emph{cursor} $c = (v,i) \in \C = \V \times \I$ is a reference to the (possibly empty) set of all vertexes with parent vertex $v$ and child index $i$. Let $c_{root} = (v_{root}, i_{root})$ be the \emph{root cursor}, a reference to all of the rooted vertexes, of a graph.

% \paragraph{Edge} \par

% %An \emph{edge} $\e = (c, v, u) \in \E = \C \times \V \times \U$ adds vertex $v$ with the set of vertexes referenced by cursor $c$.

% \paragraph{Edge State} \par

% An \emph{edge state} $s \in \S = \{\bot,+,-\}$ determines whether an edge does not exist ($\bot$), has been created (+), or has been destroyed (-).

% \paragraph{Graph Action} \par

% A \emph{graph action} $A = (\e,s,u) \in \A = \E \times \S \times \U$ is an instance of a binding from edge $\e$ to edge state $s$ that can be identified by UUID $u$. Denote by $[\e :-> s]_{u}$ a binding from edge $\e$ to edge state $s$ (identified by UUID $u$).

% \paragraph{Graph} \par

% A \emph{graph} $G = [\e :-> s]_{u}^{\ast{}} \in \G = \A^{\ast{}}$ is a function from edges to edge states. Denote by $G(\e) = s$ the state $s$ of edge $\e$ in graph $G$, and by $G A$ the extension of graph $G$ by graph action $A$.

% \subsubsection{The Editor}

% \paragraph{Move Action} \par

% %A \emph{move action} $a \in \A^{move} = \{ \Left,\Right,\Up,\Down \} \cup \{ \Select \} \times \C$ re-positions the cursor.

% \paragraph{Edit Action} \par

% An \emph{edit action} $a \in \A^{edit}= \{ \Create \} \times \K \cup \{ \Destroy \} \cup \{ \Restore \} \times \V$ adds or removes an edge at the cursor.

% \paragraph{Communication Action} \par

% A \emph{communication action} $a = (\Send, A^{\ast{}}, u^{\ast{}}) \in \A^{comm} = \{ \Send \} \times \A^{\ast{}} \times \U^{\ast{}}$ applies the graph action sequence $A^{\ast{}}$ to the editors identified by UUIDs $u^{\ast{}}$.

% \paragraph{Action} \par

% An \emph{action} $a \in \A = \A^{move} \cup \A^{edit} \cup \A^{comm}$ describes a change to one or more editors.

% \paragraph{Editor} \par

% %An \emph{editor} $E = (G,c,A_{Q}^{\ast{}},A_{H}^{\ast{}},u) \in \EE = \mathcal{G} \times \C \times \AA^{\ast{}} \times \AA^{\ast{}} \times \U$ is an instance of graph $G$ with cursor $c$, queued graph actions $A_{Q}^{\ast{}}$, and known graph action history $A_{H}^{\ast{}}$, that can be identified by UUID $u$.

% \subsubsection{The Environment} \par

% An \emph{environment} $(E^{\ast{}}, A^{\ast{}}) \in \EE^{\ast{}} \times \AA^{\ast{}}$ is a set of communicating editors $E^{\ast{}}$ that share a global graph action history $A^{\ast{}}$.

% \subsection{Sets and Relations}

% \paragraph{Edges} \par

% \begin{align*}
%   \edges(G)            & = \dom(G)                                       \\
%   \liveEdges(G)        & = \{ \e \in \edges(G) | G(\e)=+ \}              \\
%   \vertexChildren(G,v) & = \{ ((v',i),v'',u) \in \liveEdges(G) | v'=v \} \\
%   \cursorChildren(G,c) & = \{ (c',v,u) \in \liveEdges(G) | c'=c \}       \\
%   \parents(G,v)        & = \{ (c,v',u) \in \liveEdges(G) | v'=v \}
% \end{align*}

% \paragraph{Vertexes} \par

% \begin{align*}
%   \vertexes(G)        & = \{ v \in \V | \exists (c,v,u) \in \edges(G) \}              \\
%   \liveVertexes(G)    & = \{ v \in \V | \exists (c,v,u) \in \liveEdges(G) \}          \\
%   \childVertexes(G,v) & = \{ v' \in \V | \exists (c,v',u) \in \vertexChildren(G,v) \} \\
%   \orphans(G)         & = \{ v \in \liveVertexes(G) | \abs{\parents(G,v)} = 0 \}      \\
%   \multiparents(G)    & = \{ v \in \liveVertexes(G) | \abs{\parents(G,v)} > 1 \}      \\
%   \vertex(G,u)        & = v \in \vertexes(G) \iff (\exists k \in \K)(v = (k,u))
% \end{align*}

% \paragraph{Graph Actions} \par

% $\known(E_1 \cdots E_n) = A_{H1}^{\ast{}} \cdots A_{Hn}^{\ast{}}
%   \iff (\forall j=1,\ldots,n)(E_j = (G_j,c_j,A_{Qj}^{\ast{}},A_{Hj}^{\ast{}},u_j))$

% \section{Operational Semantics}

% \subsection{Movement}

% $\boxed{E,A -->a E,A}$

% \begin{mathpar}
%   \inferrule{
%   \leftIndex(i) = i'
%   }{
%   (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Left} (G,(v,i'),A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{
%   \rightIndex(i) = i'
%   }{
%   (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Right} (G,(v,i'),A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{
%   \parents(v) = \{(c',v',u')\}
%   }{
%   (G,(v,i),A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Up} (G,c',A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{
%   \cursorChildren(c) = \{(c',(k,u_k),u')\} \\
%   \downIndex(k) = i
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Down} (G,((k,u_k),i),A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }

%   \inferrule{}{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Select~c'} (G,c',A_Q^{\ast{}},A_H^{\ast{}},u),A
%   }
% \end{mathpar}

% \subsection{Editing}

% \begin{mathpar}
%   \inferrule{
%   \defaultIndex(c) = TODO(Was None) \\
%   u, u_k \in \U \text{ fresh} \\
%   A_H' = [(c,(k,u_k),u) :-> +]
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Create~k} (G A_H',c,A_Q^{\ast{}},A_H^{\ast{}}A_H',u),A
%   }

%   \inferrule{
%   \defaultIndex(c) = i' \\
%   c' = ((k,u_k),i') \\
%   u, u_k \in \U \text{ fresh} \\
%   \\\\
%   \cursorChildren(G,c) = \{(c_j,v_j,u_j')\}_{j=1}^n \\
%   A_H'^{\ast{}} =
%     [(c  ,(k,u_k),u  ) :-> +]
%     [(c' ,v_j    ,u_j') :-> +]_{j=1}^n
%   [(c_j,v_j    ,u_j') :-> -]_{j=1}^n
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Create~k}
%   (G A_H'^{\ast{}},c',A_Q^{\ast{}},A_H^{\ast{}}A_H'^{\ast{}},u),A
%   }

%   \inferrule{
%   \cursorChildren(G,c) = \{\e_j\}_{j=1}^n \\
%   A_H'^{\ast{}} = [\e_j :-> -]_{j=1}^n
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Destroy}
%   (G A_H'^{\ast{}},c,A_Q^{\ast{}},A_H^{\ast{}}A_H'^{\ast{}},u),A
%   }

%   \inferrule{
%   A_H = [(c,v,u') :-> +] \\
%   u' \text{ fresh}
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A \xrightarrow{\Restore~v} (G A_H,c,A_Q^{\ast{}},A_H^{\ast{}}A_H,u),A %
%   }
% \end{mathpar}

% \subsection{Communication}

% \paragraph{Record} \par
% %
% \begin{mathpar}
%   \inferrule{
%   a \in \A^{edit} \\
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A -->a (G',c',A_Q^{\ast{}},A_H'^{\ast{}},u),A
%   }{
%   (G,c,A_Q^{\ast{}},A_H^{\ast{}},u),A -->a (G',c',A_Q^{\ast{}}a,A_H'^{\ast{}},u),A
%   }
% \end{mathpar}

% \paragraph{Replay} \par

% $\boxed{E,A \xrightarrow{a^{\ast{}}} E,A}$
% %
% \begin{mathpar}
%   \inferrule{
%     E,A -->a E'',A'' \\
%     E'',A'' \xrightarrow{a^{\ast{}}} E',A' \\
%   }{
%     E,A \xrightarrow{aa^{\ast{}}} E',A'
%   }
% \end{mathpar}

% \paragraph{Publish} \par

% $\boxed{(E,A)^{\ast{}} -->a (E,A)^{\ast{}}}$
% %
% \begin{mathpar}
%   \inferrule{
%   E_j = (G_j,c_j,A_{Qj}^{\ast{}},A_{Hj}^{\ast{}},u_j) \\
%   E_j,A_j \xrightarrow{a'^{\ast{}}} (G_j',c_j',A_{Qj}'^{\ast{}}, A_{Hj}'^{\ast{}},u_j),A_j' \\
%   E_j' =
%   ( G_j'
%   , c_j'
%   , A_{Qj}'^{\ast{}} \setminus a'^{\ast{}}
%   , A_{Hj}'^{\ast{}} \setminus \known(E_1 \cdots E_n),u_j) \\
%   j = 1,\ldots,n
%   }{
%   (E_1,A_1) \cdots (E_n,A_n) \xrightarrow{\Send~ a'^{\ast{}} } (E_1',A_1') \cdots (E_n',A_n')
%   }
% \end{mathpar}

% Missing Action.env members: Record / Report / Stop / Replay, Dump / Load, Clone / Drop

% $G \overset{\left(e,s\right)}{\longrightarrow} G\left[ e \mapsto s \sqcup G(e) \right]$

% % Edge.Map.add action.value.edge (join old_state new_state) graph
